---
title: "Toronto Transit Commision Bus Delays: An Analysis"
subtitle: "A Journey Through a Calendar Year"
author: 
  - Xavier Ibanez-Padron
thanks: "Code and data are available at: https://github.com/Ludosss/Toronto-Bus-Delay-Analysis"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(readxl)
```


# Introduction

All throughout the world, functional transit systems have been one of the key features
of progress and development, and notable among these transit systems is the Toronto Transit Commission, or the TTC. In service since 1921 [@TTC1921], the TTC offers subway, streetcar, and bus routes throughout the City of Toronto and beyond. In particular the TTC Bus system is of interest, supplying more than half
of the TTC's entire yearly throughput, with nearly 280 million riders serviced in 2022 alone [@TTC2022]. With such a massive role in maintaining cohesion within the streets of Toronto, it is imperative to
consider not just its strengths, but also its weaknesses. Of these weaknesses, one of significant concern is delays - commonly leaving commuters stranded within the city, every day.

With 162 standard routes as of 2022 [@TTCRoutes], minimizing delays of any kind
is crucial in ensuring smooth operations. 

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

This paper uses the TTC Bus Delay Data [@busdata] retrieved from the City of Toronto's Open Data Portal via the opendatatoronto library [@datator], published by the TTC itself. This dataset was last refreshed on January 14, 2024, and contains bus delay data for the past 10 full calendar years, with every documented TTC bus delay stored within the data.

The dataset used the 2023 delay data, with over 56,000 data points. The dataset contains data regarding numerous factors that constitute a bus delay, but of particular importance for this analysis are the Date, Time, Route, Day, Delay (min), and Incident.

```{r}
#| label: cleaned_data
#| fig-cap: Table showing the cleaned data
#| echo: false

analysis_data <- read_excel(here::here("outputs/data/cleaned_bus_data.xlsx"))

kable(head(analysis_data), caption = "Sample of Cleaned TTC Bus Delay Data")
```

The process of data cleaning and analyzing was done in R [@citeR] using the tidyverse [@tverse], readxl [@rdxl], and openxlsx [@opxl] packages

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`. We also use @tidyverse


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("outputs/models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check





## Diagnostics







\newpage


# References


